% !TeX root = ../main.tex

\chapter{Introduction}

Hypervisors are essential to cloud computing. They manage the hardware
resources to provide the virtual machine (VMs) abstraction and host
these VMs in the cloud.
The widely used commodity
hypervisors, such as KVM~\cite{kivity07kvm} or Hyper-V~\cite{hyperv},
include a large and complex TCB to satisfy users' requirements in
performance and functionality. These hypervisors were written in unsafe
languages like C, making them vulnerable to safety bugs, such as
out-of-bound memory access and use-after-free. For example, KVM
integrates an entire Linux OS kernel inside its TCB. Attackers that
successfully exploit hypervisor vulnerabilities may gain the ability
to steal or modify secret VM data.

Previous work HypSec~\cite{hypsec} has retrofitted commodity hypervisors into a
small trusted core that enforces resource access control to ensure the
confidentiality and integrity of VM data against hypervisor and host operating
system exploits. However, the security of the whole system still depends on the
implementation of the small trusted TCB. Any vulnerability in the trusted TCB
can void the guarantees of VM data confidentiality and integrity.
While SeKVM~\cite{sekvm} extended the work of HypSec~\cite{hypsec} by formally
verifying the smaller TCB, the scalability of the approach is limited as formal
proofs only validate specific implementations, and as the codebase evolves to
incorporate new features or undergo code refactoring, the existing proof
becomes outdated, necessitating a new proof for any code modifications.

Rust is an emerging programming language that ensures strong memory safety
guarantees at compile time while offering performance efficiency.
Its distinctive ownership and lifetime system
effectively addresses potential safety issues that programmers may encounter.
Rust prevents various memory safety bugs, for example, null pointer
dereferences are eliminated by distinguishing between nullable and non
nullable types, nullable types are not allowed by default, array out-of-bound
accesses are prevented by runtime checks that are added by the compiler, and
Rust's ownership system prevents dangling pointers.
Further, Rust allows developers to
sacrafice the memory safety guarantees for stronger control over the program,
for example to directly manage low-level systems resources such as memory.
These low-level operations requires the \code{unsafe} keyword which marks a
region of code for which the compiler does not guarantee memory safety.
Due to these
attributes, various previous work has adopted Rust to implement systems
software with critical security and performance requirements, including
operating systems~\cite{NrOS, Redleaf, TockOS, theseus},
hypervisors~\cite{DuVisor, RustyHermit}, web browsers~\cite{servo},
and TEEs~\cite{rustsgx,rustee}.
There has been recent adoption of Rust in the mainline Linux kernel. However,
instead of replacing the existing Linux kernel code written in C with Rust,
the current efforts were limited to developing new Rust-based device drivers.

%Our work explores implementing a Linux KVM TCB in Rust, so that the resulting hypervisor
%benefits from the strong safety guarantees that Rust provides.
%Our work leverages the Rust programming language and rewrite SeKVM \cite{sekvm},
%a secure Linux KVM hypervisor in Rust, so that the resulting hypervisor
%benefits from the strong safety guarantees that Rust automatically provides.
We have developed \rustsec{} for Linux 5.15, a Rust-based secure Linux KVM hypervisor.
\rustsec{} reimplemented SeKVM \cite{sekvm} and replaced SeKVM's verified TCB
with a Rust-based TCB, called \rustcore{}.
% KRUSTVM DESCRIPTION
\rustsec{} incorporates the small Rust TCB \rustcore{} to
protect VM confidentiality and integrity against the large and untrusted
hypervisor codebase that encompasses KVM’s host Linux kernel.
% KRUSTVM DESCRIPTION END
% RUST BASED ADVANTAGES
\rustsec{} benefits from the strong safety guarantees that Rust provides.
Its Rust-based TCB also allows for frequent code updates  , since the compiler
automatically checks the safety rules for safe Rust.
This ensures that our implementation of \rustsec{} is free from memory bugs
such as null pointer dereference, use-after-free, except for places where
unsafe Rust is used.
By rewriting a C-based hypervisor to a Rust-based implementation,
most of the responsibility of human auditing is shifted to the compiler.
The manual auditing efforts can be focused solely on the
portions where unsafe Rust is utilized, rather than the entire program.

We ported the Linux 4.18-based SeKVM to Linux 5.15, a recent version of
long-term-support Linux, for \rustsec{}. This allows us to take advantage of
new kernel features, including Link-Time-Optimization (LTO) and energy-aware
scheduling.
% RUST BASED ADVANTAGES END
% CHALLENGES
During the development of \rustsec{}, we identified and overcame challenges
that arose when trying to rewrite and port SeKVM's TCB to Rust for Linux 5.15.
Firstly, the Linux kernel had undergone many changes between version 4.18 and
5.15, such as feature addition and kernel API changes. Therefore, we need to
forward port SeKVM to Linux 5.15 prior to initiating the Rust rewrite process.
Second, Linux 5.15 does not support Rust as a development language, meaning
Rust code can not be linked with the rest of the kernel by the Linux build
system. To resolve this challenge, we rolled our own Makefile and integrated
the build process of our Rust code with Linux's build system.
Third, writing a KVM TCB in a new language like Rust poses many language
compatibility issues. For example, C headers are not usable in Rust, and name
mangling exists in Rust but not in C, etc. We must address each issue  for our
implementation to work.
% CHALLENGES END
%incorporate a Rust TCB inside Linux, rewrite SeKVM's TCB in Rust, and bring up
%\rustsec{} on real hardware.
%We build upon the work of SeKVM \cite{sekvm} and forward ported SeKVM
%from Linux 4.18 to Linux 5.15, to take advantage of new kernel features
%including Link-Time-Optimization (LTO) and energy-aware scheduling.
%SeKVM's
%verified TCB \textit{\secore{}} is then rewritten in Rust, which is called
%\textit{\rustcore{}}.
%The resulting hypervisor, \rustsec{}, incorporates the small Rust TCB
%\rustcore{} to
%protect VM confidentiality and integrity against the large and untrusted
%hypervisor codebase that encompasses KVM’s host Linux kernel.
Aside from the Rust-based KVM TCB implementation challenges, we also
implemented \rustcore{} in a way such that the amount of unsafe Rust is
minimized.
Unsafe code are enclosed within a safe abstraction and a safe API is exposed
in order to implement complex functionalities in safe Rust, including CPU,
memory, VM boot protection, VM exit, and hypercall handlers.
Further, raw pointer accesses, which are unsafe in Rust, are protected using
Rust’s type system. In \rustcore{}, raw pointers are used for accessing
physical memory. Physical memory is divided into multiple disjoint regions,
and the \rustcore{} implementation guarantees that all memory accesses done by
\rustcore{} are located in the predefined regions, ensuring that bugs caused by
pointers pointing to incorrect memory regions are prevented.
This involves transforming raw pointers into references, allowing Rust to
automatically insert runtime checks for out-of-bound array indices, and
building customized Rust types that enforce bound-checking for raw pointer
accesses.
%This is done in two parts, for raw pointer accesses to the region which stores
%metadata used by \rustcore{}, called the \textit{\rustcore{} metadata region},
%these accesses are bound via a set of reference getter functions (RGF).
%Each RGF wraps a given Rcore’s raw pointer usage and returns a mutable
%reference to the associated shared metadata object after the caller acquires
%the corresponding lock. Because the raw pointer is turned into a mutable
%reference, the memory accessed is guaranteed to be bound by the size of the
%type being referenced, for arrays, the compiler automatically adds runtime
%checks that checks for out-of-bound array indices.
%For raw pointer accesses to the other regions, we built customized Rust types
%for each memory region that enforces bound-checking, and \rustcore{} accesses
%each memory region via the corresponding type.

%This is achieved with customized Rust types for each memory region that
%enforces bound-check to accesses, and mandating that \rustcore{} accesses a
%memory region via each corresponding type.

% describe the benefits
%We spent less than one person year rewriting SeKVM into \rustsec{}.
Performance evaluation of \rustsec{} on real Arm64 hardware shows that
\rustsec{} incurs modest performance overhead to application workloads
compared to mainline KVM and SeKVM. We demonstrate the practicality of
securing an existing commodity hypervisor by a C-to-Rust rewrite.

The rest of the thesis will be organized as follows. Background
will be reviewed in \autoref{sec:bg}. Our threat model and assumptions are
listed in \autoref{sec:threatmodel}. The process of implementing a Rust TCB
for KVM and the techniques used are described in \autoref{sec:rewrite}.
\autoref{sec:securercore} presents how Rust's safety features are utilized to
design and secure \rustcore{} memory accesses.
Evaluation of \rustsec{} and its comparison with mainline KVM and SeKVM is
covered in \autoref{sec:eval}. Related work and future work are discussed in
\autoref{sec:rwfw}. At last, we conclude the thesis in
\autoref{sec:conclusions}.
