\chapter{Related Work and Future Work}
\label{sec:rwfw}

\section{Related Work}

\subsection{VM Protection}
%~\cite{lisosp21,zhang2011cloudvisor,zeyu20usenix,pkvm,fidelius-hpca18,sekvm}
Various previous work redesigned the hypervisor to protect VMs.
%Hardware and software approaches have both been explored, including nested
Cloudvisor \cite{zhang2011cloudvisor, zeyu20usenix} introduced a tiny security
monitor underneath the commodity hypervisor to protect the hosted VMs.
Twinvisor \cite{lisosp21} supports regular VMs and confidential VMs by running
hypervisors within both of ARM TrustZone's normal world and secure world.
HypSec \cite{hypsec} and pkvm \cite{pkvm} reduced the hypervisor's resource access
control component into to a small core to reduce the attack surface.
Unlike our work, none of them used Rust to secure their hypervisor
implementation.
\rustsec{} and SeKVM~\cite{sekvm} both leveraged HypSec's design~\cite{hypsec}
to retrofit and secure KVM, providing the same level of VM protection. SeKVM
included a formally verified core to protect VMs against an untrusted host Linux
kernel, while \rustsec{} relies on a Rust-based \rustcore{} to protect VMs.
Formal verification of the concurrent C-based SeKVM core requires significant
effort. The authors took two person-years to complete the correctness and
security proofs.
In contrast, our Rust-based implementation took less than one person-year.
Different from formal verification, because the Rust compiler automatically
ensures memory safety properties, our hypervisor codebase is flexible to
frequent updates.

\subsection{Rust-based Systems}
Recent work extended existing C/C++ systems with a Rust binding to enable
a Rust-based programming environment. Rust-SGX~\cite{rustsgx} and RusTEE
\cite{rustee} wrapped the C/C++ TEE SDK and exposed a safe Rust API
to enable Rust programming in TEE environments such as SGX and TrustZone.
Similarly, the Rust-for-Linux~\cite{Rust-for-Linux} project added
abstraction layers to the Linux kernel to facilitate Rust driver
programming with Rust.
Besides building a Rust binding, previous work re-implemented C-based
components in virtualization systems with Rust.
rust-vmm~\cite{rust-vmm} rewrote a subset of QEMU's functionalities
and separated them into libraries in Rust crates.
Firecracker~\cite{Firecracker}, crosvm~\cite{crosvm}, Cloud Hypervisor
\cite{CloudHypervisor}, and VMSH~\cite{VMSH} extended the
rust-vmm project with extra functionalities. These previous works
built on top of existing core systems. In contrast, our work
retrofitted Linux/KVM with a Rust-based TCB.
HyperEnclave~\cite{hyperenclave} relies on a Rust-based security
monitor to enforce isolation between enclave TEEs. Unlike
our work, the authors did not discuss the Rust monitor's implementation
and its unsafe Rust usage.

\subsection{Verification and Formal Methods}

seL4 \cite{sel4} presented the first machine-checked verification of an OS
kernel. The total effort including code implementation, proof construction,
and the related research took over 20 person years.
CertiKOS \cite{certikos} presented an extensible architecture for building
certified concurrent OS kernels. Their mC2 kernel is verified to be
functionally correct and contains no bugs. Better than seL4, the mC2 kernel
is capable of running on stock x86 machines with mutliple cores. Further,
the assembly code in the kernel are also verified, and the CompCertX verified
compiler is used for compilation, these combined make the mC2 kernel certified
down to the assembly level. Overall the proof effort for certifing the mC2
kernel is 2 person years, and it consists of 6500 lines of C and x86 assembly.
SeKVM \cite{sekvm} presented a secure and formally verified Linux KVM
hypervisor. Different from seL4 and CertiKOS, which both verified less commonly
used kernels, SeKVM verified the commodity Linux KVM hypervisor, which is
widely deployed commercially. To achieve this, KVM is retrofitted into a small
trusted core \secore{} and the remaining untrusted services KServ, and
\secore{} is proved to guarantee VM confidentiality and integrity.
All of these work verified their code implementation by writing proofs to show
that the code refines, or satisfies the specifications given. The
specifications captures all the bahavior of the implementation.
However, the high level specifications themselves may be potentially insecure.
To mitigate this issue, SeKVM takes a step further and proves \secore{}'s
specification upholds VM confidentiality and integrity by showing there can not
exist any information leakage between VMs and KServ, regardless of how they
interact with \secore{}'s interface. In other words, the specifications of
\secore{} are guaranteed to exhibit the desired security properties. On the
other hand, seL4 and CertiKOS does not address this issue.

These formally verified software systems are guaranteed to be free of
programming errors, but achieving this comes at the cost of significant human
effort and reduced flexibility for making changes.
Although we hope that Rust can help us with reducing the effort needed to secure
software without sacraficing the ability to update the codebase easily,
it is challenging to solely rely on Rust to provide safety guarantees
comparable to formal verification.
Aside from the fact that there are still portions of unsafe Rust code within
Rcore that the compiler cannot ensure memory safety for,
logical errors such as setting the wrong permission bits for an NPT entry are
possible in Rust.
Rust verification tools \cite{Verus, Prusti, Creusot, Flux, aeneas} allows developers
to annotate Rust code with specifications, invariants and assertions and then
verify them formally, mathematically proving that the code satisfies the
specifications. We can utilize these tools, which offer functional correctness
guarantees, to eliminate logical errors and enhance the safety of \rustcore{}.
%Nonetheless, conducting extensive testing may be a viable option to gain
%confidence in ensuring the correct behavior of \rustcore{}.

\section{Future Work}
\label{sec:fw}

%Hardware features such as memory translation done by the Memory Management
%Unit (MMU), and exception levels, are not modeled by Rust, therefore the
%compiler is not able to check for misconfiguration or logical errors when
%building software on top of these features.
%In other words, potential logical bugs in \rustcore{} still undermines the
%security of our hypervisor.
%As an example, forgetting to clear page table entries in the host's NPT
%makes way for a compromised host kernel to read the memory of a protected VM.
%We can add one more layer of defense by taking advantage of Rust
%verification tools which base on Rust's strong type system
%\cite{Verus, Prusti, Creusot, Flux}. These tool allows developers to annotate
%Rust code with specifications, invariants and assertions and then verify them
%formally, mathematically proving that the code satisfies the specifications.

This thesis demonstrated how unsafe raw pointer accesses are protected through
isolating memory regions,
leveraging language features like automatic bound checks for array types, and
type constructor that checks for their arguments.
One aspect of this approach is that
relying solely on memory region isolation is not enough to prevent bugs caused
by raw pointers pointing to unintended addresses.
Erroneous raw pointers are not detectable by memory region isolation if they
reference an incorrect addresses within the correct region.
Basically, the granularity of the protection is limited to the size of the
memory regions. This limitation is similar to the hardware MMU, which cannot
detect violations of sub-page access permissions because its protection
granularity is at the page level.

In spite of the fact that memory region isolation alone can not guarantee the
absence of raw pointers pointing to unintended addresses,
a possible way of enhancing the protection given by memory region isolation is
to shrink the sizes of the memory regions. The smaller granularity can increase
the chances of detecting a raw pointer pointing to an unintended address.
This can be done by by using the
previously mentioned features even more extensively to enforce a separation of
memory regions that are smaller in size,
such as separating the NPT pools into the first level NPT pool,
second level NPT pool, and so on, or splitting the \rustcore{} metadata region
into multiple isolated regions, such as \code{VMInfo} region, \code{PMemInfo}
region, etc.
%However, the chances of bugs occuring can be further decreased by using the
%previous features even more extensively to achieve a more fine-grained memory region
%isolation, such as separating the NPT pools into the first level NPT pool,
%second level NPT pool, and so on, or splitting the \rustcore{} metadata region
%into multiple isolated regions, such as \code{VMInfo} region, \code{PMemInfo}
%region, etc.
%In addition, the lifetime feature of Rust remains unexplored in our work.
%In contrast to the spatial protection achieved through memory region isolation,
%there are opportunities to utilize lifetimes to enforce temporal safety
%guarantees, such as preventing the usage of a reference before another
%reference has gone out of scope. Leveraging lifetimes in this manner allows
%for the implementation of lock order enforcement, among other possibilities.
%This approach is separate from our spatial protection and holds the potential
%to further enhance the safety of our code.

%Other low-level system software may also benefit from a Rust rewrite.
%
%write other stuff in Rust as well
%  CCA
%  pkvm intel
%  other parts of the kernel
%  Rust for Linux

